---
title: "Populations, samples and statistical inference"
format:
        html:
          code-fold: false
editor_options: 
  chunk_output_type: console
bibliography: resources/bib-final.bib  
---


## A simple test of differences

- We have a two-group design and want to know if *Condition A* is different from *Condition B* **in any meaningful way**.
- To accomplish this we can test if an observed difference is very different to a reference, where any difference is up to chance
- Use the code chunk below to simulate data


```{r}
set.seed(1)
# Population 
A <- rnorm(1000, mean = 100, 10)

B <- rnorm(1000, mean = 92, 10)

# Sample
a <- sample(A, 15, replace = FALSE)
b <- sample(B, 15, replace = FALSE)

```



::: {.callout-note}

#### Group work

- Using the `t.test` function, test against the null-hypothesis of no difference between groups

- Construct a permutation test where a *reference distribution of possible outcomes* is created in a loop(!). Try to explain what the code below does.

- Calculate how many cases led to a more extreme result than the observed in your experiment.

```{r}
#| echo: true
#| eval: false
library(tidyverse)

differences <- vector()

for(i in 1:1000) {
        
        
     samp <- sample(c(a, b), 30, replace = FALSE)
        
     differences[i] <- mean(samp[1:15]) - mean(samp[16:30])
        
        
}


data.frame(differences) %>%
        ggplot(aes(differences)) + geom_histogram() + 
        geom_vline(xintercept = mean(a) - mean(b), color = "red", size = 2)


```


:::

## The effect of small and large samples.

- The sample size determines what differences we may observe.

::: {.callout-note}

#### Group work

- Re-do the experiment above with a smaller sample size, and a larger sample size.
- Report your experiment as a t-test and a permutation test.


## Limitations of p-values

- P-values are often misinterpreted!
- Find a scientific paper (full text) in any area and search for "P < 0.05" and "P > 0.05", see how authors interpret the results. 





