---
format: html
bibliography: resources/bib-final.bib
editor_options: 
  chunk_output_type: console
---


# Data wrangling and tables, part 2

See [part 1 of this workshop here](ws4-data-wrangling-tables.html).

## A file-type for reports

Tables are generated in R in specific output formats. These formats are commonly `html`, `pdf` or `docx` files. The output format type determines how a table is coded by the table generator. However, common to all outputs is the source file. Reports with tables, figures and text are built from Rmarkdown (`Rmd`) or quarto (`qmd`) files. 

The transition between Rmarkdown-files and quarto is effortless as they use the same syntax but quarto has benefits when it comes to publishing.

Quarto requires the installation of [plug-in software](https://quarto.org/docs/get-started/). The workshop is possible to follow using a Rmarkdown file also.

We will talk more about publishing reports later. For now, we just need a source file to create our table. **Start a new quarto or Rmarkdown file and save it somewhere on your computer. Preferably as part of a PROJECT**

```{r}
#| eval: false 
---
title: "Untitled"
---

        
        
        



```

## Table generators

There are at least 15 commonly used packages in R used for tables.^[See the list in the [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/table-other.html)] This really creates a jungle for the user.

This course have previously been focusing on `knitr` and the `kable` function as well as `flextable`. Both has their benefits, however, both also has drawbacks.

During this workshop we will focus on the `gt` package. This package is promising in bridging gaps in previous packages.

The `gt` package needs to be installed from CRAN.

## A basic table

We will produce a very basic table of group means at baseline from the `cyclingstudy` data set.

First we load packages and data 

```{r}
#| message: false
#| warning: false
library(gt)
library(exscidata)
library(tidyverse)


data("cyclingstudy")
```

The next step is to summarize the data of interest. We will use `VO2.max`, `sj.max` and `weight.T1` in our table, averaged over `group`

```{r}
cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        print()
```

The output from a code chunk without a table generator will display what your would typically see in the console. To format the output you will need the table generator.

We will add the generator as the last part of our pipe.


```{r}

cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt()



```

If the code is executed without compiling the source-file you will see a table in the Viewer pane in RStudio.

We proceed by formatting numbers. `gt` has many formatting functions created to make formatting of cell values easy. The function `fmt_numbers()` needs the columns to be formatted, otherwise defaults seems to give reasonable output. Be aware however of the `sep_mark = ","` option.


::: {.callout-note icon=false}

### Exercises (1)

- Find documentation for the `fmt_number` function and find out what how you may format numeric columns, including separation of thousands.
- Look up the argument `drop_trailing_zeros` in the help page for `fmt_number`.

:::


```{r}

cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt() %>%
        fmt_number(columns = VO2max:weight, sep_mark = "")



```

For our variables, VO<sub>2max</sub> may not need to have two decimal points. We can remove it from the first formatting and add another one.


```{r}
cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt() %>%
        fmt_number(columns = sj:weight) %>%
        fmt_number(columns = VO2max, 
                   decimals = 0) 
```

Column names can be added with the `cols_label` function, added as the other as part of the pipe


```{r}
cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt() %>%
        fmt_number(columns = sj:weight) %>%
        fmt_number(columns = VO2max, 
                   decimals = 0)  %>%
        cols_label(group = "Group", 
                   VO2max = md("VO<sub>2max</sub>"), 
                   sj = "Squat jump", 
                   weight = "Weight")
```

Notice that I've added the `md()` function around "VO2max" and added some additional code. This indicates with `html` tags that `2max` should be written in *subscript*.

We can also add units to each column name and print them on a different row as long as we use the `md()` function to wrap formatted text.

```{r}
cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt() %>%
        fmt_number(columns = sj:weight) %>%
        fmt_number(columns = VO2max, 
                   decimals = 0)  %>%
        cols_label(group = "Group", 
                   VO2max = md("VO<sub>2max</sub><br><small>(ml min<sup>-1</sup>)"), 
                   sj = md("Squat jump<br><small>(cm)"), 
                   weight = md("Weight<br><small>(kg)"))
```

In addition to html tags, we may use plain markdown inside the `md()` function. We will talk more about text formatting in coming workshops.


Next we will add a caption. A caption may be added to the `gt` function directly.

```{r}
cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt(caption = "Group characteristics") %>%
        fmt_number(columns = sj:weight) %>%
        fmt_number(columns = VO2max, 
                   decimals = 0)  %>%
        cols_label(group = "Group", 
                   VO2max = md("VO<sub>2max</sub><br><small>(ml min<sup>-1</sup>)"), 
                   sj = md("Squat jump<br><small>(cm)"), 
                   weight = md("Weight<br><small>(kg)"))
```

An alternative is to add a `tab_header` with a `title`. However, we can also add a caption as part of the code chunk. This will replace the caption used in `gt`. This comes with the additional benefit of being able to label tables. From the [quarto documentation](https://quarto.org/docs/authoring/cross-references.html) we can read that "for tables produced by executable code cells, include a label with a tbl- prefix to make them cross-referenceable". 

The top of the code chunk should therefore look like this

````markdown
`r ''````{r}
#| label: tbl-char
#| tbl-cap: "Group characteristics (code chunk)"
```
````

The label part must start with `tbl-` followed by a unique identifier. This identifier can be used to cross reference the table that will be auto numbered using the syntax `@tbl-char`, which will produce a link to the table: @tbl-char



```{r}
#| label: tbl-char
#| tbl-cap: "Group characteristics (code chunk)"



cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt(caption = "Group characteristics (Caption)") %>%
        fmt_number(columns = sj:weight) %>%
        fmt_number(columns = VO2max, 
                   decimals = 0)  %>%
        cols_label(group = "Group", 
                   VO2max = md("VO<sub>2max</sub><br><small>(ml min<sup>-1</sup>)"), 
                   sj = md("Squat jump<br><small>(cm)"), 
                   weight = md("Weight<br><small>(kg)")) %>%
        tab_header(title = "Group characteristics (Title)")
```


Finally we will add footnotes to the table. This is done using the `tab_footnote` function. 

```{r}
#| label: tbl-char2
#| tbl-cap: "Group characteristics (code chunk)"



cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt(caption = "Group characteristics (Caption)") %>%
        fmt_number(columns = sj:weight) %>%
        fmt_number(columns = VO2max, 
                   decimals = 0)  %>%
        cols_label(group = "Group", 
                   VO2max = md("VO<sub>2max</sub><br><small>(ml min<sup>-1</sup>)"), 
                   sj = md("Squat jump<br><small>(cm)"), 
                   weight = md("Weight<br><small>(kg)")) %>%
        tab_header(title = "Group characteristics (Title)") %>%
        tab_footnote(footnote = "Values are means")
```

We can specify a location for a reference using the locations argument.


```{r}
#| label: tbl-char3
#| tbl-cap: "Group characteristics (code chunk)"



cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        group_by(group) %>%
        summarise(VO2max = mean(VO2.max), 
                  sj = mean(sj.max),
                  weight = mean(weight.T1)) %>%
        gt(caption = "Group characteristics (Caption)") %>%
        fmt_number(columns = sj:weight) %>%
        fmt_number(columns = VO2max, 
                   decimals = 0)  %>%
        cols_label(group = "Group", 
                   VO2max = md("VO<sub>2max</sub><br><small>(ml min<sup>-1</sup>)"), 
                   sj = md("Squat jump<br><small>(cm)"), 
                   weight = md("Weight<br><small>(kg)")) %>%
        tab_header(title = "Group characteristics (Title)") %>%
        tab_footnote(footnote = "Values are means for weight", 
        locations = cells_column_labels(columns = weight))
```



## Adding more statistics

We might want to combine multiple statistics in one table, for example the mean and stamdard deviation for each variable. Below I use a different approach to summarise values per variable and group by first adding a `pivot_longer` which makes it possible to do this operation without creating multiple new variables.

A *mean-and-standard-deviation-variable* could be created by combining vectors. The mean and standard deviation is commonly presented as *mean (SD)*. Data from a column of means and a column of SD's can be combined to create a nice display using the `paste0()` function. Example:

```{r}
#| eval: false

data.frame(m = c(46.7, 47.89, 43.5),  # A vector of means
           s = c(4.21, 4.666, 3.1)) %>% # A vector of SD's
        mutate(stat = paste0(round(m, 1), 
                             " (",
                             round(s, 1), 
                             ")")) %>%
        print()

```

However, the gt package has a nice helper function that makes it possible to do this in the gt part of the pipe.


```{r}
#| label: tbl-char4
#| tbl-cap: "Group characteristics (code chunk)"



cyclingstudy %>%
        filter(timepoint == "pre") %>%
        select(subject, group, VO2.max, sj.max, weight.T1) %>%
        pivot_longer(names_to = "variable", # <1> 
                     values_to = "value", # <1>
                     cols = VO2.max:weight.T1) %>% # <1>
        group_by(group, variable) %>% # <2>
        summarise(mean = mean(value), # <2>
                  sd = sd(value)) %>% # <2>
        pivot_wider(names_from = variable, # <3>
                        values_from = c(mean, sd)) %>% # <3>
        select(group, mean_VO2.max, sd_VO2.max,  # <3>
                        mean_sj.max, sd_sj.max, # <3>
                        mean_weight.T1, sd_weight.T1) %>% # <3>
        ungroup() %>% # <4>
        
        gt(caption = "Group characteristics (Caption)") %>%
        
        fmt_number(columns = mean_VO2.max:sd_VO2.max, decimals = 0) %>% # <5>
        fmt_number(columns = mean_sj.max:sd_weight.T1, decimals = 1) %>% # <5>
         
        cols_merge(columns = c("mean_VO2.max", "sd_VO2.max"), # <6>
                   pattern = "<<{1}>> <<({2})>>") %>% # <6>
        
        cols_merge(columns = c("mean_sj.max", "sd_sj.max"), # <6> 
                   pattern = "<<{1}>> <<({2})>>") %>% # <6>
        
        cols_merge(columns = c("mean_weight.T1", "sd_weight.T1"), # <6>
                   pattern = "<<{1}>> <<({2})>>") %>% # <6>
                   

        cols_label(group = "Group", 
                   mean_VO2.max = md("VO<sub>2max</sub><br><small>(ml min<sup>-1</sup>)"), 
                   mean_sj.max = md("Squat jump<br><small>(cm)"), 
                   mean_weight.T1 = md("Weight<br><small>(kg)")) %>%
        tab_header(title = "Group characteristics (Title)") %>%
        tab_footnote(footnote = "Values are means for weight")

```

1. Using `pivot_longer` to combine multiple variables in one value/name combination.
2. Here we group by *variable* and *group* and create two new variables in a summarised data frame. This requires less code than the above example.
3. Making the table "wide" again and using select to sort the columns.
4. `gt` tables are group-sensitive. This means that if we have a grouped data frame/tibble, `gt` will use this grouping when it creates the table.
5. It is good practice to have the same number of decimals in both the mean and SD.
6. This is where we combine the columns using `column_merge`, for each new column we need a specified `column_merge`.





## Exercise/Homework: Recreating Table 1 in [@haun2018].

Data from [@haun2018] are part of the `exscidata` package as the `hypertrophy` data set. Access it by using

```{r}
#| eval: false

library(exscidata); data(hypertrophy)

glimpse(hypertrophy)

```

Try to find as many variables presented in Table 1 in the original publication as possible in the available data and format your version of the table to the best of your ability!








