---
title: "Study designs and statistical tests"
format:
        html:
          code-fold: true
editor_options: 
  chunk_output_type: console
bibliography: resources/bib-final.bib  
---


## What to analyze?

- A regression model can be used to assess the differences between treatments in a randomized controlled trial (RCT). But what do we use as the dependent variable in this model?
- Vickers [-@vickers2001] identified and compared four scenarios that might be "tested" from a comparison of two groups:

> Baseline ($b$), follow-up ($f$) and group ($g$) can be analyzed as
>
> - *Post-scores only*: $f$ is compared between $g$
> - *Percentage change*: $100 \times \frac{b-f}{b}$ is compared between $g$
> - *Change*: $b-f$ compared between $g$
> - *Analysis of co-variance (ANCOVA)*: $g$ is compared as the effect in $f=\beta_1 b + \beta_2 g$

- The choice of method should be based on its efficiency, the ability find true effects using a minimal number of participants.
- Vickers show that the use of percentage change from baseline is inefficient [@vickers2001] and could lead to violations of assumptions for statistical tests.
- The ANCOVA model (see below) is the preferred model when model assumptions holds [@vickers2001]
- We will proceed with comparison of change-scores and later discuss ANCOVA.



## Analysis of change scores 

```{r}
#| echo: true
#| message: false
#| warning: false


library(tidyverse)
data.frame(Time = c("Pre", "Post", "Pre", "Post"), outcome = c(0,1, 0,1)) %>%
        mutate(Time = factor(Time, levels = c("Pre", "Post"))) %>%
        ggplot(aes(Time, outcome)) + 
        theme_classic() + 
        ylab("Outcome") +
        theme(axis.text.y = element_blank()) +
        geom_segment(aes(x = Time, xend = 2, 
                         y = c(0.5,0.5, 0.5,1.5), 
                         yend = c(0.5,0.5, 1.5,1.5))) + 
        scale_y_continuous(limits = c(0,2)) + 
        annotate(geom = "text", x = 2.25, y = 1.5, label = "Treatment") +
        annotate(geom = "text", x = 2.25, y = 0.5, label = "Control") 


```


- In a simple analysis of two parallel groups (experimental and control), the alternative hypothesis is that the two groups are different in terms of change scores.

$$ \Delta Y_{Treatment}-\Delta Y_{Control} \neq 0 $$

- The null-hypothesis that we use in our statistical models is 

$$ \Delta Y_{Treatment}-\Delta Y_{Control} = 0 $$

- In the simple case, this equals a *t*-test.

`t.test(delta ~ group, paired = FALSE)`

- A similar scenario is a comparison between two treatments (A and B)

```{r}
#| echo: true
#| message: false
#| warning: false


data.frame(Time = c("Pre", "Post", "Pre", "Post"), outcome = c(0,1, 0,1)) %>%
        mutate(Time = factor(Time, levels = c("Pre", "Post"))) %>%
        ggplot(aes(Time, outcome)) + 
        theme_classic() + 
        ylab("Outcome") +
        theme(axis.text.y = element_blank()) +
        geom_segment(aes(x = Time, xend = 2, 
                         y = c(0.5,0.5, 0.5,1.5), 
                         yend = c(1.1,0.5, 1.5,1.5))) + 
        scale_y_continuous(limits = c(0,2)) + 
        annotate(geom = "text", x = 2.25, y = 1.5, label = "A") +
        annotate(geom = "text", x = 2.25, y = 1.1, label = "B") 

```

- The null-hypothesis is that

$$ \Delta Y_{A}-\Delta Y_{B} = 0 $$

- The `tenthirty` data set has similar data as in the above scenario
- We want to know what treatment to recommend for 1RM improvements. 


:::{.callout-Practice}

#### Group work: Ten-thirty

- Using the `tenthirty` data, use filter to keep time-point "pre" and "post" for the leg-press exercise.
- Make a plot of the averages of each group.
- Calculate a change score as `post` - `pre`.
- Make a plot of the average change score for each group.
- Fit a regression model (`change ~ group`) and interpret the results.



### Questions:

1. Concerning the sample: Which group has the largest improvement i leg-press 1RM? 
2. Concerning inference to the population: Which method do you recommend for improving 1RM leg-press?

:::


```{r}
#| echo: true
#| eval: false
#| message: false
#| warning: false
#| code-summary: "Possible solutions"


## Create a plot of averages over time ----------- ##

library(tidyverse); library(exscidata)
data("tenthirty")

# Save the data set in a filtered version
tenthirty_reduced <- tenthirty %>%
  filter(time %in% c("pre", "post"), 
         exercise == "legpress", 
         !is.na(load)) %>% 
  print()

# Create a line plot with one line per group
tenthirty_reduced %>%  
  group_by(time, group) %>%
  summarise(m = mean(load)) %>%
  mutate(time = factor(time, levels = c("pre", "post"))) %>%
  ggplot(aes(time, m, color = group, 
             group = group)) + geom_line()


# Create a plot of average changes per group
tenthirty_reduced %>%
  pivot_wider(names_from = time, 
              values_from = load) %>%
  mutate(change = post - pre) %>%
  group_by(group) %>%
  summarise(mean_change = mean(change, na.rm = TRUE)) %>%
  ggplot(aes(group, mean_change, fill = group)) + 
  geom_bar(stat = "identity", width = 0.3)
  

# Calculate the average change per group, and difference between groups
tenthirty_reduced %>%
  pivot_wider(names_from = time, 
              values_from = load) %>%
  mutate(change = post - pre) %>%
  group_by(group) %>%
  summarise(mean_change = mean(change, na.rm = TRUE)) %>%
  pivot_wider(names_from = group, 
              values_from = mean_change) %>%
  mutate(diff = RM30 - RM10) %>%
  print()


# Calculate change score and save new data set
dat <- tenthirty_reduced %>%
  pivot_wider(names_from = time, 
              values_from = load) %>%
  mutate(change = post - pre) %>%
  print()


# Fit model with change scores explained by group
m1 <- lm(change ~ group, data = dat)

# Get the model summary
summary(m1)



```



```{r}
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Group averages over time"
library(exscidata)

tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  mutate(time = factor(time, levels = c("pre", "post"))) %>%
  group_by(time, group) %>%
  summarise(m = mean(load, na.rm = TRUE)) %>%
  ggplot(aes(time, m, group = group, color = group)) + geom_line() +
  labs(x = "Time", y = "Legpress performance (kg)", color = "Group") +
          theme_classic()
```



## From t-test to ANCOVA

- Above we used the simple comparison of change score. A more appropriate model to study outcomes of a RCT is an analysis of co-variance (ANCOVA)
- In an ANCOVA, change-scores or post values are compared and the estimates tells you the adjusted differences between groups if they had the exact same starting point.
- The ANCOVA model can be written as

$$f =\beta_0 + \beta_1\times b + \beta_2\times g $$
where $f$ is the follow-up score, $b$ is the baseline and $g$ is the group.

- In R code, this corresponds to 

`post ~ pre + group`


:::{.callout-Practice}

#### Group work: Ten-thirty ANCOVA

- Fit an ANCOVA model with change scores as the dependent variable (`m1`)
- Fit an ANCOVA model with the post values as the dependent variable (`m2`)
- Pre-training values and group should be covariates in both models.
- Use the `aov` command to fit an ANOVA model (`m3`)
- Evaluate models using `summary` and `anova`

### Questions

1. Concerning inference to the population: Which method do you recommend for improving 1RM leg-press?
2. In what ways are the models different?
3. What is happening with the `aov` command, how is it different to `m2`?

:::


```{r}
#| echo: true
#| eval: false
#| message: false
#| warning: false
#| code-summary: "Possible solutions"


# Create a data set with change

dat <- tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  pivot_wider(names_from = time, 
              values_from = load) %>%
  mutate(change = post - pre, 
         pre = pre - mean(pre)) %>% # Mean center pre-values
  print()


# Fit models 

m1 <- lm(post ~ pre + group, data = dat)

m2 <- lm(change ~ pre + group, data = dat)


summary(m1)
summary(m2)


m3 <- aov(post ~ pre + group, data = dat)

summary(m3)
anova(m1)


```


## Understanding the ANCOVA model

- The ANCOVA estimates the difference between e.g. post scores given a (estimated) association between pre- and post-scores.
- This can be visualized in a simple plot where the difference in the paralell lines are the treatment effect


```{r}
#| echo: true
#| message: false
#| warning: false


m1 <- tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  pivot_wider(names_from = "time", values_from = "load") %>%
  mutate(change = post - pre) %>%
  lm(post ~ pre + group, data = .) 


tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
    pivot_wider(names_from = "time", values_from = "load") %>%
  ggplot(aes(pre, post, color = group)) + geom_point() + 
  scale_color_manual(values = c("red", "blue")) +
  geom_abline(intercept = coef(m1)[1], slope = coef(m1)[2], color = "red")  + 
  geom_abline(intercept = coef(m1)[1] + coef(m1)[3], slope = coef(m1)[2], color = "blue") + 
  labs(title = "Estimates from an ANCOVA model", 
       subtitle = "post ~ pre + group", 
       color = "Group") + theme_classic()

```


- By correcting for baseline values we can expect different estimates for each group, notice also that the standard errors are changing

```{r}
#| echo: true
#| message: false
#| warning: false
#| fig-cap: "Group estimates from two different models, error bars are standard errors (SE)" 

library(emmeans)

m0 <- tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  pivot_wider(names_from = "time", values_from = "load") %>%
  mutate(change = post - pre) %>%
  lm(post ~ group, data = .) 

m2 <- tenthirty %>%
  filter(exercise == "legpress", 
         time %in% c("pre", "post")) %>%
  pivot_wider(names_from = "time", values_from = "load") %>%
  mutate(change = post - pre) %>%
  lm(change ~ group, data = .) 





bind_rows(data.frame(emmeans(m0, specs = ~ group)) %>%
        mutate(Model = "Post only"), 
        data.frame(emmeans(m1, specs = ~ group)) %>%
        mutate(Model = "ANCOVA")) %>%
        ggplot(aes(group, emmean, fill = Model)) + 
        geom_errorbar(aes(ymin = emmean, ymax = emmean + SE), 
                      position = position_dodge(), width = 0.5) +
        geom_bar(stat = "identity", position = position_dodge(), 
                 width = 0.5)  
        

```

- Calculating the differences between groups we can also see differences, here we also incorporate the change score model.

```{r}
#| fig-cap: "Group estimates from two different models, error bars are 95% confidence intervals." 
#| echo: true
#| message: false
#| warning: false

bind_rows(
data.frame(confint(contrast(emmeans(m0, specs = ~ group)), adjust = "none")) %>%
        mutate(Model = "Post only"),
data.frame(confint(contrast(emmeans(m1, specs = ~ group)), adjust = "none")) %>%
        mutate(Model = "ANCOVA"),
data.frame(confint(contrast(emmeans(m2, specs = ~ group)), adjust = "none")) %>%
        mutate(Model = "Change score")) %>%
        filter(contrast == "RM10 effect") %>%
        ggplot(aes(Model, estimate, fill = Model)) + 
        geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
        geom_point(shape = 21, size = 3) + 
        labs(y = "Estimated mean difference between groups") + 
        geom_hline(yintercept = 0, lty = 2)







```

- The change-score model will have lower power compared to the ANCOVA model when the correlation between pre- and post-scores decrease [@vickers2001]










