---
title: "Writing reports and working with R projects"
---


These workshop notes contains links to relevant documentation-


## Quarto and Rmarkdown

- Quarto and R Markdown a special kind of scripts where text and computer code can be combined to generate reports. 
- Under the hood, a report generator is converting code and plain text to an output format such as *html*, *pdf* or *docx* (more formats are available).
- Quarto is a new, [well documented](https://quarto.org/) format that gives extra flexibility, but also requires installation of extra software.
- R Markdown is even more well documented, see e.g. [R Markdown](https://rmarkdown.rstudio.com/), [R Markdown: The Definitev Guide](https://bookdown.org/yihui/rmarkdown/)



## Code execution and the environment

- When a quarto or Rmarkdown file is "knitted", the source file looks for e.g., data files in the same directory as the source file is saved.
- Working in a RStudio projects makes it easy to work with the report interactively as you can use relative paths.
- [RStudio has an excellent guide to its project feature](https://support.rstudio.com/hc/en-us/articles/200526207-Using-RStudio-Projects)


## RStudio projects

- A projects is basically a collection of settings together with a root directory.
- Settings can be accessed in Tools > Project Options.
- This means that you will be able to work with relative paths. If reading a `csv` file using relative paths, your code will look like this from a project.

```{r}
#| eval: false

dat <- read_csv("./data/my-data.csv")
```


- If your are using absolute paths, reaching the same goal could look like this

```{r}
#| eval: false

dat <- read_csv("C:/Users/Daniel/Dropbox/Some-folder/a-project/data/my-data.csv")

```

- RStudio projects helps you create good habits for reproducible analysis as all analyses are conducted within a stand-alone folder structure. Your data and scripts can be shared.

- Use a basic structure for all projects:

```{}
My-project
        |
        |-.Rproj        (The project settings)
        |--/data        (Contains all data needed for your analysis)
        |--/R           (Contains all scripts/R-files)
        |--/output      (Collection of all output files)


```

- Start a new project from the Project menu. 


## Writing in Quarto/R Markdown

- The basic syntax in quarto/R Markdown files is markdown. Markdown makes it easy to format text without point-and-click as all formatting can be added with syntax, example:

```{}
This text is an example of the markdown syntax which includes **bold**, *italic*,
^super^ and ~subscript~ and ~~striketrough~~
```

Resulting in:

> This text is an example of the markdown syntax which includes **bold**, *italic*,
^super^ and ~subscript~ and ~~striketrough~~

- In addition to text formatting, the markdown syntax offers solutions for adding images, tables, equations, lists, diagrams, different highligt blocks. See the [the quarto documentation](https://quarto.org/docs/authoring/markdown-basics.html) for more information.


## Code chunks

- Code chunks are sections of your source file containing code. We primarily write R-code, but e.g., python is also possible.
- The code chunk comes with several options specified in the top of the chunk using `#|`, such as:

````markdown
`r ''````{r}
#| eval: true
#| echo: true
#| warning: false
#| error: true
#| include: false
```
````
```
```

Note that the `#|` is a "new" intervention, some documentation will still suggest that you use code chunk settings in the code chunk header.

## Inline code

- Code may be included inline to include code generated outputs

```{r}
#| eval: false
a_variable <- 3.14

```

- To include the variable in the text:

```{r}
#| eval: false
The variable will be displayed here `r a_variable` 
```

## Bibliographies

- Bibliographies/Citations may be added to reports using a external bibliography file. A simple format is bibtex
- Pubmed entries can be searched using [TexMed](https://www.bioinformatics.org/texmed/)
- In the visual editor, bibliographies can be easily created

## Some notes on different formats

- HTML is the basic output from R Markdown and quarto. This is suitable for first drafts.
- To be able to render PDF files you must have an installation of rendering software. TinyTeX is generally recomended, [see the documentation and installation instructions here](https://yihui.org/tinytex/)
- Word documents creates an editable document with associated pros and cons.

- All formats has different advantages and offer flexibility that gives opportunities to create any type of document

- Other formats such as presentations, webpages, apps etc. makes quarto / R Markdown very versatile. 








